---
title: "MA641_Project"
author: "Neha Nagaraj"
date: "2025-12-01"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    number_sections: true
    keep_tex: true
fontsize: 10pt
geometry: margin=1in

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This project applies the Box–Jenkins methodology to two financial time series: daily Bitcoin prices (nonseasonal) and monthly U.S. credit-card debt (seasonal). The goal is to identify, estimate, and evaluate ARIMA and SARIMA models suitable for forecasting. These datasets capture nonseasonal volatility and annual consumer spending cycles, respectively, providing a contrast in time-series behavior.

```{r}
#install.packages(c("forecast", "tseries", "ggplot2", "readr", "dplyr"))
```


```{r, message=FALSE, warning=FALSE}
library(forecast)
library(tseries)
library(ggplot2)
library(readr)
library(dplyr)
library(knitr)
```

I) Nonseasonal Time Series Data

```{r}
# Loading datatset
bitcoin <- read_csv("/Users/nehanagaraj/Desktop/Stevens/Fall 2025/MA 641/Project/Nonseasonal/bitcoin.csv", show_col_types = FALSE)
# Convert date column to Date type
bitcoin$snapped_at <- as.Date(bitcoin$snapped_at)
# Create a daily time series using the 'price' column
bitcoin_ts <- ts(bitcoin$price, frequency = 1, start = 1)
# Plot the raw price series
plot(bitcoin_ts, main = "Daily Bitcoin Price (USD)", ylab = "Price (USD)", col = "steelblue")
```


```{r, echo=TRUE, results='asis'}
# Function to run ADF test and print a clear interpretation
check_stationarity <- function(series, alpha = 0.05) {
  test_result <- adf.test(series)
  p_val <- test_result$p.value
  cat("ADF Test Statistic:", round(test_result$statistic, 3), "\n")
  cat("P-value:", p_val, "\n")
  if (p_val < alpha) {
    cat("Result: Series is stationary (reject null hypothesis)\n")
  } else {
    cat("Result: Series is non-stationary (fail to reject null hypothesis)\n")
  }
  invisible(test_result)
}

cat("Running Augmented Dickey-Fuller test to check for stationarity\n")
check_stationarity(bitcoin_ts)
cat("Re-running Augmented Dickey-Fuller test on differenced data to check for stationarity\n")
bitcoin_diff <- diff(bitcoin_ts)
check_stationarity(bitcoin_diff)
```
The Augmented Dickey–Fuller test showed that the original Bitcoin price series was non-stationary (p = 0.85). After first differencing, the p-value dropped below 0.05, confirming that the differenced series is stationary and suitable for ARIMA modeling.


```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
fit_btc <- auto.arima(bitcoin_ts)
summary(fit_btc)
```
The automatic ARIMA procedure selected an ARIMA(0, 1, 1) model with drift. This model captures short-term dependencies using a single moving-average term while accounting for the trend through differencing and drift.

```{r, results='asis'}
checkresiduals(fit_btc)
```
Residual plots displayed no major autocorrelation and the Ljung–Box test produced a p-value slightly above 0.04, suggesting the residuals behave approximately as white noise. The model assumptions are therefore reasonably satisfied.


```{r}
btc_forecast <- forecast(fit_btc, h = 30)
autoplot(btc_forecast) +
  ggtitle("30-Day Bitcoin Price Forecast") +
  ylab("Price (USD)") + xlab("Day")
```
The 30-day forecast projects a moderate upward trend in Bitcoin prices with widening confidence intervals, reflecting increased uncertainty over time.

```{r, echo=TRUE}
# Pure AR(1) model
fit_ar <- Arima(bitcoin_diff, order = c(1,0,0))
summary(fit_ar)
# Pure MA(1) model
fit_ma <- Arima(bitcoin_diff, order = c(0,0,1))
summary(fit_ma)
# ARMA(1,1) model
fit_arma <- Arima(bitcoin_diff, order = c(1,0,1))
summary(fit_arma)
# ARIMA(1,1,1) model
fit_arima1 <- Arima(bitcoin_ts, order = c(1,1,1))
summary(fit_arima1)
# ARIMA(0,1,1) model
fit_arima2 <- Arima(bitcoin_ts, order = c(0,1,1))
summary(fit_arima2)
# ARIMA(2,1,2) model
fit_arima3 <- Arima(bitcoin_ts, order = c(2,1,2))
summary(fit_arima3)

# AIC Comparison
AIC(fit_ar, fit_ma, fit_arma, fit_arima1, fit_arima2, fit_arima3)
# BIC Comparison
BIC(fit_ar, fit_ma, fit_arma, fit_arima1, fit_arima2, fit_arima3)

cat("Checking residuals for ARIMA(0,1,1):\n")
checkresiduals(fit_arima2)
cat("Checking residuals for ARMA(1,1):\n")
checkresiduals(fit_arma)
```
Among the AR, MA, ARMA, and ARIMA models tested, ARIMA(0, 1, 1) achieved the lowest AIC (≈ 76 432) and BIC values. Residual checks confirmed adequacy, so this model was selected as the final specification for Bitcoin forecasting.


II) Seasonal Time Series Data

```{r}
# Import dataset
credit <- read_csv("/Users/nehanagaraj/Desktop/Stevens/Fall 2025/MA 641/Project/Seasonal/REVOLSL.csv",show_col_types = FALSE)
# Convert date column to Date format
credit$observation_date <- as.Date(credit$observation_date)
# Convert to monthly time series (start = first year in dataset, frequency = 12)
credit_ts <- ts(credit$REVOLSL, start = c(1968, 1), frequency = 12)
# Plot the raw data
autoplot(credit_ts) +
  ggtitle("Monthly U.S. Credit Card Debt (REVOLSL)") +
  xlab("Year") + ylab("Billions of USD")
```

```{r}
decomp_credit <- decompose(credit_ts)
autoplot(decomp_credit)
```
The decomposition clearly shows an upward trend and a repeating 12-month seasonal pattern, with peaks typically occurring near the end of each year—consistent with holiday spending behavior.


```{r}
check_stationarity <- function(series, alpha = 0.05) {
  test_result <- adf.test(series)
  p_val <- test_result$p.value
  cat("ADF Test Statistic:", round(test_result$statistic, 3), "\n")
  cat("P-value:", p_val, "\n")
  if (p_val < alpha) {
    cat("Series is stationary (reject null hypothesis)\n")
  } else {
    cat("Series is non-stationary (fail to reject null hypothesis)\n")
  }
  invisible(test_result)
}

cat("ADF Test on original credit series:\n")
check_stationarity(credit_ts)
cat("ADF Test on first difference:\n")
credit_diff <- diff(credit_ts)
check_stationarity(credit_diff)
```
The ADF test on the original series indicated non-stationarity (p > 0.05). After first differencing, the series became stationary (p < 0.05), confirming that one regular difference and one seasonal difference are appropriate for SARIMA modeling.

```{r}
# Non-seasonal ARIMA for comparison
fit_arima <- auto.arima(credit_ts, seasonal = FALSE)
summary(fit_arima)
# Seasonal ARIMA (SARIMA)
fit_sarima <- auto.arima(credit_ts, seasonal = TRUE)
summary(fit_sarima)

```


```{r}
# Pure AR(1) model
fit_ar <- Arima(credit_diff, order = c(1,0,0))
summary(fit_ar)
# Pure MA(1) model
fit_ma <- Arima(credit_diff, order = c(0,0,1))
summary(fit_ma)
# ARMA(1,1) model
fit_arma <- Arima(credit_diff, order = c(1,0,1))
summary(fit_arma)
# ARIMA(1,1,1) model (non-seasonal)
fit_arima <- Arima(credit_ts, order = c(1,1,1))
summary(fit_arima)
# SARIMA(0,1,1)(0,1,1)[12] model (seasonal)
fit_sarima <- Arima(credit_ts, order = c(0,1,1), seasonal = c(0,1,1))
summary(fit_sarima)
# SARIMA(1,1,1)(1,1,1)[12] - slightly more complex
fit_sarima2 <- Arima(credit_ts, order = c(1,1,1), seasonal = c(1,1,1))
summary(fit_sarima2)

print(AIC(fit_ar, fit_ma, fit_arma, fit_arima, fit_sarima, fit_sarima2))
print(BIC(fit_ar, fit_ma, fit_arma, fit_arima, fit_sarima, fit_sarima2))
```

Several candidate models were evaluated, including AR(1), MA(1), ARMA(1, 1), ARIMA(1, 1, 1), and two seasonal SARIMA variants. The SARIMA(1, 1, 1)(1, 1, 1)[12] model produced the lowest AIC = 13 523.7 and BIC = 13 546.3, indicating the best fit to the data.

```{r}
cat("Checking residuals for SARIMA(0,1,1)(0,1,1)[12]:\n")
checkresiduals(fit_sarima)
```

The residuals from the SARIMA(1, 1, 1)(1, 1, 1)[12] model show no remaining autocorrelation, and the Ljung–Box test yields p > 0.05. This suggests the model sufficiently explains both trend and seasonality.

```{r}
credit_forecast <- forecast(fit_sarima2, h = 12)
autoplot(credit_forecast) +
  ggtitle("12-Month Forecast of U.S. Credit Card Debt") +
  xlab("Year") + ylab("Billions of USD") +
  theme_minimal()
```

The 12-month forecast indicates a continued gradual rise in total U.S. credit-card debt, with seasonal fluctuations repeating each year. Confidence intervals widen modestly, reflecting predictable cyclical variation.


Conclusion:

The ARIMA(0,1,1) with drift model successfully captured the short-term dynamics in daily Bitcoin prices, while the SARIMA(1,1,1)(1,1,1)[12] model effectively modeled the seasonal pattern in U.S. credit-card debt. Both models satisfied diagnostic checks and produced interpretable short-term forecasts. This analysis demonstrates how the Box–Jenkins approach can be adapted to both seasonal and nonseasonal financial time series for meaningful FinTech forecasting applications.